---
title: "Population momentum"
author: "Monica Alexander and Robert Chung"
output: html_document
---

```{r setup, include=FALSE}
# run this chunk first!
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

In this lab we are going to explore the concept of population momentum. We'll be using the same data set from yesterday, which had fertility, mortality and population data for six countries (Australia, Kenya, Senegal, Russia, Singapore and Niger). 

First read in the data and define the Leslie matrix function, the same as before. 

```{r}
d <- read.csv("./projection_data_long.csv")
# alter the nLx and nFx columns so we dont have to worry about units later
d <- d %>% 
  mutate(nLx = nLx/10^5, nFx = nFx/1000)

## function to define Leslie matrix, based on nLx and nFx values
leslie <- function(nLx,
                   nFx, 
                   n_age_groups=10,
                   ffab = 0.4886){
  L = matrix(0, nrow = n_age_groups, ncol = n_age_groups)
  L[1,] = ffab * nLx[1]*(nFx[1:n_age_groups]+nFx[2:(n_age_groups+1)]*nLx[2:(n_age_groups+1)]/nLx[1:n_age_groups])/2 # top row 
  diag(L[2:n_age_groups,1:(n_age_groups-1)]) = nLx[2:n_age_groups] / nLx[1:(n_age_groups-1)] # subdiagonal
  return(L)
}

# define ffab 
ffab <- 0.4886
```

Define a function `pop_project` that does a population projection and outputs the result. The function takes as inputs 

- A leslie matrix
- A vector of initial population values
- The number of age groups (we set our default as 10, for five year age groups 0-50)
- The number of projection steps. 

```{r}
# make a projection function, based on code in Lab 1
pop_project <- function(A, # Leslie matrix
                        initial_pop, 
                        n_age_groups = 10,
                        n_projections = 50){
  
  K <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
  K[,1] <- initial_pop[1:n_age_groups]
  
  # do the projection!
  for(i in 2:(n_projections+1)){
    K[,i] <- A%*%K[,i-1] 
  }
  
  # get K in a form that's easier to ggplot
  # make into a dataframe, name columns as years and add an age column
  Kdf <- as.data.frame(K)
  colnames(Kdf) <- seq(from = 2015, to = (2015+n_projections*5), by = 5)
  Kdf <- cbind(age = seq(from = 0, to = 45, by = 5), Kdf)
  
  # get in long format and then add proportion of population in each age group
  Klong <- Kdf %>% 
    gather(year, population, -age) %>%
    group_by(year) %>%
    mutate(proportion = population/sum(population), 
           age = as.factor(age)) %>%
    ungroup()
  
  return(Klong)
}

```

Let's check to see this function works okay by doing a population projection for Kenya. 

```{r}

# get nLx and nFx values from our dataset
nLx.kenya <- d$nLx[d$country=="Kenya"]
nFx.kenya <- d$nFx[d$country=="Kenya"]

A <- leslie(nLx = nLx.kenya, 
              nFx = nFx.kenya)
inital_pop <- d$nKx[d$country=="Kenya"]

project_kenya <- pop_project(A, inital_pop)
tail(project_kenya)

```

Plot the total population over time: 

```{r}
# have a look at Kenya's population over time
tot_pop <- project_kenya %>% 
  group_by(year) %>%
  summarise(pop = sum(population)) %>%
  mutate(R = c(NA, log(pop[2:n()]/pop[1:(n()-1)])/5))

ggplot(data = tot_pop, aes(x = year, y = pop, group = 1)) +
  geom_point() + geom_line()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle("Kenya - total population over time")


```

```{r}
r.kenya <- log(Re(eigen(A)$values[1]))/5 # stable growth rate
v <- Re(eigen(A)$vectors[,1])
k.kenya <- v/sum(v) # stable age population

r.kenya
k.kenya
```

```{r}
ggplot(d, aes(x = age, y = nFx, group= country, color = country)) + geom_line()
```


## What happens if you change the fertility rates (=quantum)? 
We can shift the fertility curve of Kenya up or down. For example, if we double the the growth rate, what do you expect to happen to the:

- Total population size
- Stable growth rate
- Stable age population


```{r}
nFx.kenya.double <- nFx.kenya*2
nFx.kenya.double
```

```{r}
A.double <- leslie(nLx = nLx.kenya, 
              nFx = nFx.kenya.double)

project_kenya_double <- pop_project(A.double, inital_pop)
tail(project_kenya_double)
```

```{r}
# have a look at Kenya's population over time
tot_pop_double <- project_kenya_double %>% 
  group_by(year) %>%
  summarise(pop = sum(population)) %>%
  mutate(R = c(NA, log(pop[2:n()]/pop[1:(n()-1)])/5))

ggplot(data = tot_pop_double, aes(x = year, y = pop, group = 1)) +
  geom_point() + geom_line()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle("Kenya - total population over time - double fertility rates")
```


```{r}
r.kenya.double <- log(Re(eigen(A.double)$values[1]))/5 # stable growth rate
v <- Re(eigen(A.double)$vectors[,1])
k.kenya.double <- v/sum(v) # stable age population

r.kenya.double
k.kenya.double
```

Compare the doubled fertility rates with the initial fertility rates -- age structure is younger. 

```{r}
cbind(k.kenya, k.kenya.double)
```


## What happens if you change the timing (=tempo)?

Now pretend that the fertility age schedule shifts to the right - so the $_nF_x$ values stay the same, but are shifted one age group to the right. What do you expect to happen to the:

- Total population size
- Stable growth rate
- Stable age population

Use the same code as the quantum example above, but just change your input $_nF_x$ values.

```{r}
length(nFx.kenya)
```

```{r}
nFx.kenya.shift <- c(0, nFx.kenya[-21])
nFx.kenya.shift
```
```{r}
A.shift <- leslie(nLx = nLx.kenya, 
              nFx = nFx.kenya.shift)

project_kenya_shift <- pop_project(A.shift, inital_pop)
tail(project_kenya_shift)
```

```{r}
# have a look at Kenya's population over time
tot_pop_shift <- project_kenya_shift %>% 
  group_by(year) %>%
  summarise(pop = sum(population)) %>%
  mutate(R = c(NA, log(pop[2:n()]/pop[1:(n()-1)])/5))

ggplot(data = tot_pop_shift, aes(x = year, y = pop, group = 1)) +
  geom_point() + geom_line()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle("Kenya - total population over time - shifted fertility rates")
```


```{r}
r.kenya.shift <- log(Re(eigen(A.shift)$values[1]))/5 # stable growth rate
v <- Re(eigen(A.shift)$vectors[,1])
k.kenya.shift <- v/sum(v) # stable age population

r.kenya.shift
k.kenya.shift
```
Compare the initial fertility rates with the shifted fertility rates -- age structure is older. 

```{r}
cbind(k.kenya, k.kenya.shift)
```


## Keyfitz scenario

What happens if we suddenly decrease all fertility rates so that the overall NRR is equal to 1 (and the long run population is stationary). To do this we need to calculate the NRR:

```{r}
# get fertility measures 
nrr_df <- d %>% 
  group_by(country) %>% 
  summarise(nrr = sum(nFx*nLx*ffab))
nrr_df
```

Now calculate some new $_nF_x$ values that are divided through by the NRR:

```{r}
# divide all fertility rates through by NRR
d <- d %>% 
  left_join(nrr_df, by = 'country') %>%
  mutate(nFx_switch = nFx/nrr)
head(d)
```

```{r}
# convince ourselves that new NRRs = 1
d %>% 
  group_by(country) %>% 
  summarise(nrr_switch = sum(nFx_switch*nLx*ffab))

```

Repeat the kenya projection using the new $_nF_x$ values:

```{r}
# repeat kenya projection with new nFx values
A_keyfitz <- leslie(nLx = d$nLx[d$country=="Kenya"], 
              nFx = d$nFx_switch[d$country=="Kenya"])
inital_pop <- d$nKx[d$country=="Kenya"]

project_kenya_keyfitz <- pop_project(A_keyfitz, inital_pop)

# plot total population over time
tot_pop_keyfitz <- project_kenya_keyfitz %>% 
  group_by(year) %>%
  summarise(pop = sum(population)) %>%
  mutate(R = c(NA, log(pop[2:n()]/pop[1:(n()-1)])/5))

ggplot(data = tot_pop_keyfitz, aes(x = year, y = pop, group = 1)) +
  geom_point() + geom_line()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle("Kenya - total population over time with NRR=1")


```

```{r}
# according to the stationary population identity, b*e0 should equal one

# get final year of projection
final_year <- project_kenya_keyfitz$year[nrow(project_kenya_keyfitz)]

# get final population by age in final year
final_pop <- project_kenya_keyfitz %>% 
  filter(year == final_year[[1]]) %>%
  select(population) 

# get nFx values after the switch
nFx_switch <- d %>% 
  filter(country=="Kenya", age < 50) %>% # filter out ages above 50
  select(nFx_switch)

# crude birth rate in final year
b_final <- sum(nFx_switch*final_pop)/sum(final_pop)*ffab

# life expectancy
e0 <- d %>% 
  filter(country=="Kenya", age<50) %>%
  summarise(sum(nLx))

# this should equal 1
b_final*e0[[1]]

```
We can get an idea of how much past age structure affects the ultimate population size by looking at the ratio of the first to last populations:

```{r}
## momentum factor 

K_pre_drop <- tot_pop_keyfitz$pop[1]
K_final <- tot_pop_keyfitz$pop[nrow(tot_pop_keyfitz)]
K_final/K_pre_drop

```
